\documentclass[12pt]{article}
\usepackage{graphicx}

\begin{document}

\title{Reverse Engineering CAN 101}
\author{Bhargava Manja}
\maketitle

First, I will explain how CAN works for different parts of the car in different
regimes to help you reason about CAN, then detail my techniques for reverse
engineering.

\section{CAN}
\subsection{Introduction}
Electronic control units (ECUs) control most functions of modern vehicles. They
are small embedded computers that all talk to one another with a controller
area network (CAN), sort of like a LAN. The difference between LAN networking
and CAN networking is that all CAN devices are connected to all other CAN devices
because they are all connected to a common CAN bus. ECUs communicate over this
bus via CAN frames, the CAN analogue to an ethernet packet. If every ECU can
hear all the talk over the CAN bus, how does an ECU know which frames are meant
for it to listen and respond to? Each CAN frame begins with an id that uniquely
identifies the ECU that is meant to receive the frame. Each frame also has
a body, which contains data for the recipient. 

\subsection{Frames}
There are two kinds of frames: standard and extended. The only difference
between the two is that extended frames have longer ids and more room for data.
All ECUs on a bus must be using the same kind of frame. I have wasted hours not
being able to communicate over a bus before realizing I am sending extended
frames on a standard CAN network.

\subsection{Frequency}
ECUs are configured to communicate over the CAN bus via a specific frequency.
That is, if I have one CAN bus with two ECUs, they must be configured to
communicate with the same frequency. You can think of frequency like language
in that two ECUs on two different frequencies can't understand one another.

\section{Grigore's car}
Grigore's car actually contains 2 CAN networks. One is the body CAN, or BCAN,
over which most components of the car communicate. For example, the headlights
and windshield wiper ECUs are on the BCAN. They work like so:
\begin{enumerate}
\item when a driver turns the blinker on, they change the state on a control
   ECU under the steering wheel
\item the control ECU sends a stream of messages to the headlight ECU. These 
   messages each say the same thing, which amounts to "turn left blinker on" 
   (for example). This sort of message is a \textbf{stream message}. If only one is
   sent, the blinker will only turn on for a fraction of a second. The control
   ECU must continue to send messages for the blinker to continue to blink.
\item If the driver turns off the blinker and turns on high beams, the control ECU
   ends the first stream and
   sends another stream to the same headlight ECU, this time with a different
   data field indicating a different action (high beams vs blinkers).
\end{enumerate}
The BCAN operates at 125 kbps, and uses extended frames.

The other network is the fast CAN, or FCAN. FCAN connects components that can
not be obstructed, and must operate quickly. For example, FCAN connects your
pedal to the engine. FCAN on Grigore's car operates at 500 kbps and uses
standard frames. 

\subsection{Atomic Messages}
The door open and close functionality does not use stream messages, but uses
\textbf{atomic messages}. That is, one message is sufficient to open or close the
door. There may be other such exceptions, but this is the only one so far.
The door open/close buttons on the driver dash are actually wired straight to
the motors of the car and don't use CAN. The key open/close buttons send
a radio signal to an ECU that always stays on and sends a single open/close
message to the door ECU

\section{Reverse Engineering}
To reverse engineer a function of the car, you need to find out \textbf{the ID of the
ECU that controls that function} and the \textbf{data payload to trigger that function}. 
Then you can send a frame over the bus with the proper ID and payload to
activate that function. To determine both of these, I do the following:
\begin{enumerate}
\item Connect the USB-CANII tool to a CAN splice
\item Start the CANDebug software tool and record the target vehicle function's
CAN trace
\item Run the FrameDecoder tool to find candidate ECU ids for further
inspection
\item Use CANDebug's \textit{display filter} and \textit{summary display}
capabilities to winnow down the list of possible ECU ids
\end{enumerate}

I assume you can do number 1, and will explain the rest below:
\subsection{Recording Traces via CANDebug}
First, click "Save Real Time" and choose a file name. Then click "Start" and 
choose the appropriate parameter for baudrate, and click "Ok".

Perform the target function (e.g turn the windshield wipers on or clock the
door lock button on the key). Through some trail and error I have determined
the following heuristic best practices:
\begin{itemize}
\item Do not collect target logs for longer than 10 seconds. I usually stick to
between 3 and 5
\item record multiple target logs
\item Before each recording, I turn the car off, turn it on (ignition), wait
for a few seconds, and begin recording
\end{itemize}

\subsection{Finding Candidate ECUs}
Before you continue, open up your logs and delete the first 3 lines of each.
They are gunk that will cause the FrameDecoder tool to fail. The help text of
FrameDecoder:
\begin{verbatim}
  -h, --help            show this help message and exit
  -c, --code-gen        generate C code for STM32 boards
  -i IDLE_LOG, --idle-log=IDLE_LOG
                        contains frames generated when car is idling
 -a ACTION_LOG, --action-log=ACTION_LOG
                        contains frames generated when car does action we want
                        to reverse engineer
  -r IGNORED_IDS, --ignore=IGNORED_IDS
                        comma separated list of message IDs toignore when
                        generating diff
\end{verbatim}
This tool takes an \textit{idle log} (a CAN trace of the car idling) and an
\textit{action log} (a log of the target action taking place, what you have
just recorded) and returns a list of CAN frames that are in the action log but
are not in the idle log, in the order in which they occur. This is a candidate
list of ECUs that might control the target function. 

Now, we must pare this list down. Remember, what we want is the specific
ECU id and payload that activates a particular function of the vehicle. Let us
assume that we have some way, in CANDebug, to filter log output by CAN id. That
is, we can set CANDebug to only show messages to particular ids. Then, we could do the
target action manually (say, by turning on the car's headlights) and see if any
messages to our filtered ids show up. We can discard the ids that don't show
up, and focus on those that do. Fortunately this is not a hypothetical
scenario - CANDebug includes an option to filter log messages by CAN id! The
"Open" menu contains a "Set display filter" option which allows you to specify
CAN ids, and clicking "enable display filter" in the main toolbar will only
show messages to those ids. 

Of the CAN ids that survive the initial filtering process, only one will be the correct
CAN id, so the other messages to other ids are unrelated, tangentially related,
or random. Successive iterations of this filter and log loop will make it
obvious which messages are random and unrelated. The harder ids to eliminate
are those with function related to the target function. What we can  do now
that we have a smaller set of candidate CAN ids is to take each candidate id
and payload and \textit{send} it over the vehicle's CAN bus somehow. Messages
that have no effect can be discarded. Again, CANDebug provides a "Send CAN
frame" dialog that allows you to send standard and extended frames with
arbitrary ids and payloads, and has a host of options for cycling messages,
sleeping between messages, etc. Here you must be \textbf{careful}: it is
possible to throw away the correct CAN id and payload thinking they have no
effect. Test each one carefully, trying out various send parameters. Try
sending the same message multiple times in a row with varying inter-message
pauses, or cycle the message. Be very sure it does not have the desired effect
before discarding it. 

CANDebug's \textit{summary display} feature may be of help here. Normally
CANDebug will display each received message on a new line, making it difficult
to correlate action to log message. By clocking "CAN 1 Summary Display" in the
open menu, you can change the UI to better facilitate the correlation between
CAN message and action. In the summary display, each unique id in your filter
has its own line, and displays that id's last received message, flashing when
a new message to that id is sent (read by USB-CANII). 

\end{document}
